---
title: "Sports Analytics Project"
author: "Jack Arbuckle, Ian Pezzella, and Reuben Dayal"
date: "2023-10-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load Packages:
library(tidyverse)
library(ggplot2) # Load ggplot2
library(ggdark) # Load ggdark
library(ggimage) # Load ggimage
library(reshape) # Load reshape
library(ggridges) # Load ggridges
library(ggrepel) # Load ggrepellibrary(tidyverse) # Load tidyverse
library(GGally) # Load GGally

```

```{r}
# Load Functions:
source("sports_analytics_functions.r")
#source("pbp_functions.r")
```


```{r}
load("team_logos.rda")
```


```{r}
library(cfbfastR)
# Load data
cfb_2023 <- load_cfb_pbp(seasons = 2023, weeks = c(1, 2, 3,4,5,6,7,8,9,10,11))
class(cfb_2023)
# Convert to data frame
cfb_2023 <- as.data.frame(cfb_2023)
```


```{r}
# Create summarized play type
s_play_type <- create_simplex_play_type(cfb_2023)
# Add summarized play type back to dataset
cfb_2023$playtype <- s_play_type
```


```{r}
# Create team level summary
team_summary <- create_team_summary(pbp = cfb_2023, # Set dataset
                                team_logos = team_logos) # Set logos to 
```

The first few columns relate to team level information


* team_id - The ID code for each team
* school - The school name
* color - The school primary color
* alt_color - The school alternative color
* logos - Link to the school logo image
* games - The number of games played so far

The next few columns relate to the offensive performance of the school:

* pass_n - Number of passes thrown
* pass_yards - Total pass yards 
* pass_epa - Total pass EPA (Expected points added)
* pass_wpa - Total pass WPA (Win probability added)

* rush_n - Number of rushes
* rush_yards - Total rush yards
* rush_epa - Total rush EPA (Expected points added)
* rush_wpa - Total Rush WPA (Win probability added)

We then have the offensive statistics at a per play level:

* pass_yards_per_play - Pass yards per play    
* pass_epa_per_play - Pass EPA (Expected points added) per play
* pass_wpa_per_play - Pass WPA (Win probability added) per play

* rush_yards_per_play - Rush yards_per_play
* rush_epa_per_play - Rush EPA (Expected points added) per play
* rush_wpa_per_play - Rush WPA (Win probability added) per play     

The next set of columns relate to the defensive performance of the school:

* pass_n_def - Number of passes defended
* pass_yards_def - Total pass yards against
* pass_epa_def - Total pass EPA against (Expected points added)
* pass_wpa_def - Total pass WPA against (Win probability added)

* rush_n_def - Number of rushes defended
* rush_yards_def - Total rush yards against
* rush_epa_def - Total rush EPA against (Expected points added)
* rush_wpa_def - Total rush WPA against (Win probability added)
 
Finally we have the defensive statistics at a per play level:

* pass_yards_per_play_def - Pass yards against per play
* pass_epa_per_play_def - Pass EPA against (Expected points added) per play
* pass_wpa_per_play_def - Pass WPA against (Win probability added) per play

* rush_yards_per_play_def - Rush yards against per play
* rush_epa_per_play_def - Rush EPA against (Expected points added) per play
* rush_wpa_per_play_def - Rush WPA against (Win probability added) per play


```{r}
CFP_schools <- team_summary %>%
  filter(school %in% c('Washington', 'Ohio State', 'Michigan','Florida State', 'Georgia'))

CFP_schools
```
Power Five Opponents
```{r}
wash_opponents <- team_summary %>%
  filter(school %in% c('Michigan State', 'Colorado', 'Arizona', 'Oregon','Arizona State','Stanford', 'USC', 'Utah'))
wash_opponents

osu_opponents <- team_summary %>%
  filter(school %in% c('Indiana', 'Notre Dame', 'Maryland', 'Purdue', 'Penn State', 'Wisconsin', 'Rutgers', 'Michigan State'))
osu_opponents

uga_opponents <- team_summary %>%
  filter(school %in% c('South Carolina', 'Auburn', 'Kentucky', 'Vanderbilt', 'Florida','Missouri', 'Ole Miss'))
uga_opponents

fsu_opponents <- team_summary %>%
  filter(school %in% c('Louisiana', 'Boston College', 'Clemson','Virginia Tech', 'Syracuse', 'Duke', 'Wake Forest', 'Pittsburgh','Miami'))
fsu_opponents

mich_opponents <- team_summary %>%
  filter(school %in% c('Rutgers', 'Nebraska', 'Minnesota','Indiana', 'Michigan State', 'Purdue', 'Penn State'))
mich_opponents
```


```{r}
average_fsu<- c((mean(fsu_opponents$pass_epa_per_play)),(mean(fsu_opponents$pass_epa_per_play_def)), (mean(fsu_opponents$rush_epa_per_play_def)), (mean(fsu_opponents$rush_epa_per_play)))

average_mich<- c((mean(mich_opponents$pass_epa_per_play)),(mean(mich_opponents$pass_epa_per_play_def)), (mean(mich_opponents$rush_epa_per_play_def)),(mean(mich_opponents$rush_epa_per_play)))

average_osu<- c((mean(osu_opponents$pass_epa_per_play)),(mean(osu_opponents$pass_epa_per_play_def)), (mean(osu_opponents$rush_epa_per_play_def)), (mean(osu_opponents$rush_epa_per_play)))

average_uga<- c((mean(uga_opponents$pass_epa_per_play)),(mean(uga_opponents$pass_epa_per_play_def)), (mean(uga_opponents$rush_epa_per_play_def)), (mean(uga_opponents$rush_epa_per_play)))

average_wash<- c((mean(wash_opponents$pass_epa_per_play)),(mean(wash_opponents$pass_epa_per_play_def)), (mean(wash_opponents$rush_epa_per_play_def)), (mean(wash_opponents$rush_epa_per_play)))


all_opponents_avg <- data.frame(
  school = c("Florida State", "Michigan", "Washington", "Ohio State", "Georgia"),
  opp_pass_epa_per_play = c(average_fsu[1], average_mich[1], average_wash[1], average_osu[1], average_uga[1]),
  opp_pass_epa_per_play_def = c(average_fsu[2], average_mich[2], average_wash[2], average_osu[2], average_uga[2]),
  opp_rush_epa_per_play_def = c(average_fsu[3], average_mich[3], average_wash[3], average_osu[3], average_uga[3]),
  opp_rush_epa_per_play = c(average_fsu[4], average_mich[4], average_wash[4], average_osu[4], average_uga[4]))
  
all_opponents_avg <- merge(all_opponents_avg, CFP_schools, by = "school")




# Display the resulting data frame
print(all_opponents_avg)
```



```{r}
# Create plot
OPP_Def_EPA_Per_Play_plot <- ggplot(data = all_opponents_avg, # Set dataset
                 aes(x = opp_pass_epa_per_play_def, # Set x-axis variable
                     y = opp_rush_epa_per_play_def)) + # Set y-axis variable
  geom_point() + # Set geom point to create scatter plot
  geom_image(image = CFP_schools$logos, asp = 16/9) + # Add logos
   labs(x = "Opponents Pass EPA DEF Per Play", # Add labels
       y = "Opponents Rush EPA Def Per Play",
       title = "Opponents: Rush v Pass EPA Def Per Play",
       subtitle = "Higher EPA indicates team has played weaker defenses") 

# Turn off dark mode
invert_geom_defaults()


ggsave(OPP_Def_EPA_Per_Play_plot, file = 'OPP_Def_EPA_Per_Play_plot.jpg', width = 10, height = 10, dpi = 600)
```

```{r}
# Create plot
OPP_Off_EPA_Per_Play_plot <- ggplot(data = all_opponents_avg, # Set dataset
                 aes(x = opp_pass_epa_per_play, # Set x-axis variable
                     y = opp_rush_epa_per_play)) + # Set y-axis variable
  geom_point() + # Set geom point to create scatter plot
  geom_image(image = CFP_schools$logos, asp = 16/9) + # Add logos
   labs(x = "Opponent Pass EPA Per Play", # Add labels
       y = "Opponent Rush EPA Def Per Play",
       title = "Opponents: Rush v Pass EPA Per Play",
       subtitle = "Higher EPA indicates team has played stronger offenses")

# Turn off dark mode
invert_geom_defaults()


ggsave(OPP_Off_EPA_Per_Play_plot, file = 'OPP_Off_EPA_Per_Play_plot.jpg', width = 10, height = 10, dpi = 600)
```


```{r}
# Create plot
Off_EPA_Per_Play_plot <- ggplot(data = CFP_schools, # Set dataset
                 aes(x = rush_epa_per_play, # Set x-axis variable
                     y = pass_epa_per_play)) + # Set y-axis variable
  geom_point() + # Set geom point to create scatter plot
  geom_image(image = CFP_schools$logos, asp = 16/9) + # Add logos
  dark_theme_bw() + # Set dark theme
   labs(x = "Rush EPA Per Play", # Add labels
       y = "Pass EPA Per Play",
       title = "Rush v Pass EPA Per Play",
       subtitle = "Higher EPA indicates stronger offense") 

# Turn off dark mode
invert_geom_defaults()

ggsave(Off_EPA_Per_Play_plot, file = 'Off_EPA_Per_Play_plot.jpg', width = 10, height = 10, dpi = 600)
```



```{r}
# Create plot
Def_EPA_Per_Play_plot <- ggplot(data = CFP_schools, # Set dataset
                 aes(x = rush_epa_per_play_def, # Set x-axis variable
                     y = pass_epa_per_play_def)) + # Set y-axis variable
  geom_point() + # Set geom point to create scatter plot
  geom_image(image = CFP_schools$logos, asp = 16/9) + # Add logos
  dark_theme_bw() + # Set dark theme
   labs(x = "Rush EPA DEF Per Play", # Add labels
       y = "Pass EPA DEF Per Play",
       title = "Rush v Pass EPA DEF Per Play",
       subtitle = "Lower EPA indicates stronger defense") 

# Turn off dark mode
invert_geom_defaults()


ggsave(Def_EPA_Per_Play_plot, file = 'Def_EPA_Per_Play_plot.jpg', width = 10, height = 10, dpi = 600)
```

```{r}
number_of_passvsrush_plays_plot <- ggplot(data = CFP_schools, # Set dataset
                 aes(x = rush_n, # Set x-axis variable
                     y = pass_n)) + # Set y-axis variable
  geom_point() + # Set geom point to create scatter plot
  geom_image(image = CFP_schools$logos, asp = 16/9) + # Add logos
  dark_theme_bw() + # Set dark theme
   labs(x = "Number of Rushes", # Add labels
       y = "Number of Passes",
       title = "Rush v Passing Plays",
       subtitle = "CFB 2023") 

# Turn off dark mode
invert_geom_defaults()


ggsave(number_of_passvsrush_plays_plot, file = 'number_of_passvsrush_plays_plot.jpg', width = 10, height = 10, dpi = 600)
```

Create weighted EPA for all teams based on Schedule
```{r}
weighted_off_def <- all_opponents_avg %>%
  mutate(weighted_passing_off = 10*((1/(0.5+opp_pass_epa_per_play_def))*(pass_epa_per_play+0.5)))%>%
mutate(weighted_rushing_off = 10*((1/(0.5+opp_rush_epa_per_play_def)) * (rush_epa_per_play+0.5)))%>%
  mutate(weighted_passing_def = 10*((1/(0.5+opp_pass_epa_per_play)) * (pass_epa_per_play_def+0.5)))%>%
  mutate(weighted_rushing_def = 10*((1/(0.5+opp_rush_epa_per_play)) * (rush_epa_per_play_def+0.5)))

weighted_off_def
```



```{r}
weighted_off_plot<- ggplot(data = weighted_off_def, # Set dataset
                 aes(x = weighted_off_def$weighted_rushing_off, # Set x-axis variable
                     y = weighted_off_def$weighted_passing_off)) + # Set y-axis variable
  geom_point() + # Set geom point to create scatter plot
  geom_image(image = CFP_schools$logos, asp = 16/9) + # Add logos
  dark_theme_bw() + # Set dark theme
   labs(x = "Weighted Rushing Offense", # Add labels
       y = "Weighted Passing Offense",
       title = "Weighted Offense",
       subtitle = "CFB 2023") 

# Turn off dark mode
invert_geom_defaults()

ggsave(weighted_off_plot, file = 'Weighted_Off_Plot.jpg', width = 10, height = 10, dpi = 600)
```



```{r}

weighted_def_plot <- ggplot(data = weighted_off_def, # Set dataset
                 aes(x = weighted_off_def$weighted_rushing_def, # Set x-axis variable
                     y = weighted_off_def$weighted_passing_def)) + # Set y-axis variable
  geom_point() + # Set geom point to create scatter plot
  geom_image(image = CFP_schools$logos, asp = 16/9) + # Add logos
  dark_theme_bw() + # Set dark theme
   labs(x = "Weighted Rushing Defense", # Add labels
       y = "Weighted Passing Defense",
       title = "Weighted Defense",
       subtitle = "CFB 2023") 

# Turn off dark mode
invert_geom_defaults()


ggsave(weighted_def_plot, file = 'Weighted_Def_Plot.jpg', width = 10, height = 10, dpi = 600)
```



```{r}
weighted_pass_defvsoff_plot <- ggplot(data = weighted_off_def, # Set dataset
                 aes(x = weighted_off_def$weighted_passing_def, # Set x-axis variable
                     y = weighted_off_def$weighted_passing_off)) + # Set y-axis variable
  geom_point() + # Set geom point to create scatter plot
  geom_image(image = CFP_schools$logos, asp = 16/9) + # Add logos
  dark_theme_bw() + # Set dark theme
   labs(x = "Weighted Passing Defense", # Add labels
       y = "Weighted Passing Offense",
       title = "Passing Def vs Passing Off",
       subtitle = "CFB 2023") 

# Turn off dark mode
invert_geom_defaults()


ggsave(weighted_pass_defvsoff_plot, file = 'weighted_pass_defvsoff.jpg', width = 10, height = 10, dpi = 600)
```



```{r}
weighted_rush_defvsoff_plot <- ggplot(data = weighted_off_def, # Set dataset
                 aes(x = weighted_off_def$weighted_rushing_def, # Set x-axis variable
                     y = weighted_off_def$weighted_rushing_off)) + # Set y-axis variable
  geom_point() + # Set geom point to create scatter plot
  geom_image(image = CFP_schools$logos, asp = 16/9) + # Add logos
  dark_theme_bw() + # Set dark theme
   labs(x = "Weighted Rushing Defense", # Add labels
       y = "Weighted Rushing Offense",
       title = "Rushing Def vs Rushing Off",
       subtitle = "CFB 2023") 

# Turn off dark mode
invert_geom_defaults()


ggsave(weighted_rush_defvsoff_plot, file = 'weighted_rush_defvsoff.jpg', width = 10, height = 10, dpi = 600)
```

```{r}
washvsosu <- weighted_off_def %>%
  filter(school %in% c("Washington", "Ohio State"))%>%
mutate(Wash_Passing_Off = 10*(weighted_passing_off[2] * weighted_passing_def[1]))%>%
mutate(Wash_Rushing_Off = 10*(weighted_rushing_off[2] * weighted_rushing_def[1]))%>%
mutate(OSU_Passing_Off = 10*(weighted_passing_off[1] * weighted_passing_def[2]))%>%
mutate(OSU_Rushing_Off = 10*(weighted_rushing_off[1] * weighted_rushing_def[2]))

```


```{r}
selected_columns <- c("Wash_Passing_Off", "Wash_Rushing_Off", "OSU_Passing_Off", "OSU_Rushing_Off")

# Melt the dataframe to long format for easier plotting
melted_data_osu <- melt(washvsosu, id.vars = "school", measure.vars = selected_columns)

selected_rows_osu <- melted_data_osu %>%
  slice(c(2, 4, 5, 7))

# Create a grouped bar plot
washvsosu_plot<- ggplot(selected_rows_osu, aes(x = variable, y = value,fill = school)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Offense: Washington vs. Ohio State",
       x = "Offense Rating", y = "Values") +
  scale_fill_manual(values = c("firebrick1", "purple")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

washvsosu_plot

ggsave(washvsosu_plot, file = 'washvsosu_plot.jpg', width = 10, height = 10, dpi = 600)

```
```{r}
washvsmich <- weighted_off_def %>%
  filter(school %in% c("Washington", "Michigan"))%>%
mutate(Wash_Passing_Off = 10*(weighted_passing_off[2] * weighted_passing_def[1]))%>%
mutate(Wash_Rushing_Off = 10*(weighted_rushing_off[2] * weighted_rushing_def[1]))%>%
mutate(Mich_Passing_Off = 10*(weighted_passing_off[1] * weighted_passing_def[2]))%>%
mutate(Mich_Rushing_Off = 10*(weighted_rushing_off[1] * weighted_rushing_def[2]))

selected_columns <- c("Wash_Passing_Off", "Wash_Rushing_Off", "Mich_Passing_Off", "Mich_Rushing_Off")

# Melt the dataframe to long format for easier plotting
melted_data_mich <- melt(washvsmich, id.vars = "school", measure.vars = selected_columns)

selected_rows_mich <- melted_data_mich %>%
  slice(c(2, 4, 5, 7))

# Create a grouped bar plot
washvsmich_plot <- ggplot(selected_rows_mich, aes(x = variable, y = value, fill = school)) +
  geom_bar(stat = "identity", position = "dodge") +
  dark_theme_bw() + # Set dark theme
  labs(title = "Comparison of Offense: Washington vs. Michigan",
       x = "Offense Rating", y = "Values") +
  scale_fill_manual(values = c("navy", "purple")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

washvsmich_plot

ggsave(washvsmich_plot, file = 'washvsmich_plot.jpg', width = 10, height = 10, dpi = 600)
```

```{r}
washvsfsu <- weighted_off_def %>%
  filter(school %in% c("Washington", "Florida State"))%>%
mutate(Wash_Passing_Off = 10*(weighted_passing_off[2] * weighted_passing_def[1]))%>%
mutate(Wash_Rushing_Off = 10*(weighted_rushing_off[2] *weighted_rushing_def[1]))%>%
mutate(FSU_Passing_Off = 10*(weighted_passing_off[1] * weighted_passing_def[2]))%>%
mutate(FSU_Rushing_Off = 10*(weighted_rushing_off[1] * weighted_rushing_def[2]))

selected_columns <- c("Wash_Passing_Off", "Wash_Rushing_Off", "FSU_Passing_Off", "FSU_Rushing_Off")

# Melt the dataframe to long format for easier plotting
melted_data_fsu <- melt(washvsfsu, id.vars = "school", measure.vars = selected_columns)

selected_rows_fsu <- melted_data_fsu %>%
  slice(c(2, 4, 5, 7))

# Create a grouped bar plot
washvsfsu_plot<- ggplot(selected_rows_fsu, aes(x = variable, y = value, fill = school)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Offense: Washington vs. Florida State",
       x = "Offense Rating", y = "Values") +
  scale_fill_manual(values = c("darkred", "purple")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

washvsfsu_plot

ggsave(washvsfsu_plot, file = 'washvsfsu_plot.jpg', width = 10, height = 10, dpi = 600)
```


```{r}
washvsuga <- weighted_off_def %>%
  filter(school %in% c("Washington", "Georgia"))%>%
mutate(Wash_Passing_Off = 10*(weighted_passing_off[2] * weighted_passing_def[1]))%>%
mutate(Wash_Rushing_Off = 10*(weighted_rushing_off[2] * weighted_rushing_def[1]))%>%
mutate(UGA_Passing_Off = 10*(weighted_passing_off[1] * weighted_passing_def[2]))%>%
mutate(UGA_Rushing_Off = 10*(weighted_rushing_off[1] * weighted_rushing_def[2]))

selected_columns <- c("Wash_Passing_Off", "Wash_Rushing_Off", "UGA_Passing_Off", "UGA_Rushing_Off")

# Melt the dataframe to long format for easier plotting
melted_data_uga <- melt(washvsuga, id.vars = "school", measure.vars = selected_columns)

selected_rows_uga <- melted_data_uga %>%
  slice(c(2, 4, 5, 7))

# Create a grouped bar plot
washvsuga_plot <- ggplot(selected_rows_uga, aes(x = variable, y = value, fill = school)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Offense: Washington vs. Georgia",
       x = "Offense Rating") +
  scale_fill_manual(values = c("red", "purple")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

washvsuga_plot

ggsave(washvsuga_plot, file = 'washvsuga_plot.jpg', width = 10, height = 10, dpi = 600)
```



